Type theories as quotient inductive-recursive types

### 2025-02-21 tex/main.tex
Group constructors by sort instead of by feature in Appendix

\paragraph*{Quotient inductive-inductive-recursive types}
We give the QIIRTs for type theories with a universe of small types, $\Pi$-types, and extensional identity types in one place below.

\subparagraph*{Substitution calculus with a base type $\UU$}
\begin{alignat*}{3}
  \Ctx      & : && \Set                   \\
  \Ty       & : && \Ctx \to \Set          \\
  \Sub      & : && \Ctx \to \Ctx \to \Set \\
  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
\begin{alignat*}{5}
  \emptyctx & : && \Ctx \\
  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
\begin{alignat*}{5}
\subTy{\blank}{\blank} & : \implicit{\Gamma, \Delta}\; && \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \\
  \UU     & : \implicit{\Gamma}\; && \Ty\; \Gamma
\end{alignat*}
\begin{alignat*}{5}
  \emptysub & : \implicit{\Gamma} && \Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma}) \to \Sub\;\Gamma\,(\Delta, A) \\
  \idS & : \implicit{\Gamma} && \Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta} && \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
  \pi_1 & : \implicit{\Gamma, \Delta, A}\; && \Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\subTy{A}{\pi_1\;\sigma}) \\
  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A} && \Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma})
\end{alignat*}
\begin{alignat*}{3}
  & \subTy{A}{\idS}           && \reduce A \\
  & \subTy{A}{\sigma ; \tau}  && \reduce \subTy{\subTy{A}{\tau}}{\sigma} \\
  & \subTy{A}{\pi_1(\sigma, t)} && \reduce \subTy{A}{\sigma} \\
  & \subTy{A}{\pi_1(\sigma; \tau)} && \reduce \subTy{\subTy{A}{\pi_1\tau}}{\sigma} \\
  & \subTy{\UU}{\sigma} && \reduce \UU
\end{alignat*}
\begin{alignat*}{5}
  ;\text{-}\mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
  ;\text{-}\mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , \sub{t}{\sigma} \\
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\;\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\;\Gamma\; A} &&  t \\
  \emptyctx\eta   & : \implicit{\Gamma, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  [\idS]         & : \implicit{\Gamma, A, t}\; && {\sub{t}{\idS}}          && =^{\Tm\;\Gamma\;A} && t \\
  [;]            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {\sub{t}{\sigma ; \tau}} && =^{\Tm\;\Gamma\;\subTy{A}{\sigma ; \tau}} && {\sub{(\sub{t}{\tau})}{\sigma}}
\end{alignat*}

\subparagraph*{Universe of small types}
\begin{alignat*}{5}
  \subTm{\blank}{\blank}        & : \implicit{\Gamma, \Delta, A}\; \Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to  \Tm\;\Gamma\;(\subTy{A}{\sigma}) \\
 \subTm{t}{\idS}                & \reduce t \\
 \subTm{t}{\sigma ; \tau}       & \reduce \subTm{\subTm{t}{\tau}}{\sigma} \\
 \subTm{t}{\pi_1(\sigma, s)}    & \reduce \subTm{t}{\sigma} \\
 \subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{\subTm{t}{\pi_1\tau}}{\sigma} \\
 \subTm{t}{\sigma}              & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
\end{alignat*}
%
\begin{alignat*}{5}
  \El       & : \implicit{\Gamma}\; && \Tm\;\Gamma\;U \to \Ty\;\Gamma
\end{alignat*}
%
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} \reduce \El\,(\subTm{u}{\sigma})
\end{alignat*}

\subparagraph*{$\Pi$-types}
\begin{alignat*}{5}
  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
  \sigma^+ & \defeq (\pi_1\idS; \sigma), \pi_2\sigma
\end{alignat*}
\begin{alignat*}{5}
  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
\idS                \uparrow A  & \reduce \idS \\
\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\tau}) ; (\tau \uparrow A) \\
\pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
\pi_1(\sigma; \tau) \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwise}
\end{alignat*}
\begin{alignat*}{5}
  \Pi             & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
  \lambda         &: \implicit{\Gamma, A, B} \; && \Tm\;(\Gamma, A)\;B \to \Tm\;\Gamma(\Pi\,A\,B) \\
  \mathsf{app}    &: \implicit{\Gamma, A, B} \; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
\begin{alignat*}{3}
  & \subTy{(\Pi\,A\,B)}{\sigma}       && {} \reduce {} && \Pi\,(\subTy{A}{\sigma})\,(\subTy{B}{\sigma\uparrow A})
\end{alignat*}
\begin{alignat*}{5}
  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
  \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\,(\Gamma, A)\,B} && t \\
  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\,\Gamma\,(\Pi\,A\,B)} && \lambda(\mathsf{app}\,t)
\end{alignat*}

\subparagraph*{Extensional identity types}

\begin{alignat*}{5}
  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) \to \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\El\,a) \to \Ty\;\Gamma
\end{alignat*}
\begin{alignat*}{5}
  \subTy{(\mathsf{Id}\;a\;t\;u)}{\sigma} \reduce \mathsf{Id}\,(\subTm{a}{\sigma})\,(\subTm{t}{\sigma})\,(\subTm{u}{\sigma})
\end{alignat*}
\begin{alignat*}{5}
  \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a)) \to \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,t) \\
  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \sub{(\mathsf{refl}\;t)}{\sigma} =^{\Tm\;\Gamma\;(\subTy{(\mathsf{Id}\,a\,t\,t)}{\sigma})} \mathsf{refl}(\sub{t}{\sigma}) \\
  \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,u) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
\end{alignat*}

>>>>

\paragraph*{Type theory as a quotient inductive-inductive-recursive type}
We give the QIIRT for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.

\begin{alignat*}{3}
  \Ctx      & : && \Set                   \\
  \Ty       & : && \Ctx \to \Set          \\
  \Sub      & : && \Ctx \to \Ctx \to \Set \\
  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
\begin{alignat*}{5}
  \emptyctx     & : \Ctx \\
  \blank,\blank & : (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
\begin{alignat*}{5}
\subTy{\blank}{\blank} & : \implicit{\Gamma, \Delta}\; && \Ty\;\Delta \to \Sub\;\Gamma\;\Delta \to \Ty\;\Gamma \\
  \UU         & : \implicit{\Gamma}\; && \Ty\; \Gamma \\
  \El         & : \implicit{\Gamma}\; && \Tm\;\Gamma\;U \to \Ty\;\Gamma \\
  \Pi         & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
  \mathsf{Id} & : \implicit{\Gamma}\; && (a : \Tm\;\Gamma\;\UU) \to \Tm\;\Gamma\;(\El\,a) \to \Tm\;\Gamma\;(\El\,a) \to \Ty\;\Gamma
\end{alignat*}

\begin{alignat*}{5}
  \emptysub & : \implicit{\Gamma} && \Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma}) \to \Sub\;\Gamma\,(\Delta, A) \\
  \idS & : \implicit{\Gamma} && \Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta} && \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
  \pi_1 & : \implicit{\Gamma, \Delta, A}\; && \Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta
\end{alignat*}

\begin{alignat*}{5}
  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A} && \Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\subTy{A}{\sigma}) \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\subTy{A}{\pi_1\;\sigma}) \\
  \lambda         &: \implicit{\Gamma, A, B} \; && \Tm\;(\Gamma, A)\;B \to \Tm\;\Gamma(\Pi\,A\,B) \\
  \mathsf{app}    &: \implicit{\Gamma, A, B} \; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B \\
  \mathsf{refl} & : \implicit{\Gamma, a}\;&&(t : \Tm\;\Gamma\;(\El\,a)) \to \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,t)
\end{alignat*}

\begin{alignat*}{5}
  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
  \sigma^+ & \defeq (\pi_1\idS; \sigma), \pi_2\sigma
\end{alignat*}

\begin{alignat*}{5}
  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, \subTy{A}{\sigma})\;(\Delta, A) \\
\idS                \uparrow A  & \reduce \idS \\
\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\tau}) ; (\tau \uparrow A) \\
\pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
\pi_1(\sigma; \tau) \uparrow A  & \reduce (\sigma \uparrow \subTy{A}{\pi_1 \tau}) ; (\pi_1 \tau \uparrow A) \\
\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwise}
\end{alignat*}

\begin{alignat*}{5}
  \subTm{\blank}{\blank}        & : \implicit{\Gamma, \Delta, A}\; \Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to  \Tm\;\Gamma\;(\subTy{A}{\sigma}) \\
 \subTm{t}{\idS}                & \reduce t \\
 \subTm{t}{\sigma ; \tau}       & \reduce \subTm{\subTm{t}{\tau}}{\sigma} \\
 \subTm{t}{\pi_1(\sigma, s)}    & \reduce \subTm{t}{\sigma} \\
 \subTm{t}{\pi_1(\sigma; \tau)} & \reduce \subTm{\subTm{t}{\pi_1\tau}}{\sigma} \\
 \subTm{t}{\sigma}              & \reduce \sub{t}{\sigma}, \quad \text{otherwise}
\end{alignat*}

\begin{alignat*}{3}
  \subTy{A}{\idS}                & \reduce A \\
  \subTy{A}{\sigma ; \tau}       & \reduce \subTy{\subTy{A}{\tau}}{\sigma} \\
  \subTy{A}{\pi_1(\sigma, t)}    & \reduce \subTy{A}{\sigma} \\
  \subTy{A}{\pi_1(\sigma; \tau)} & \reduce \subTy{\subTy{A}{\pi_1\tau}}{\sigma} \\
  \subTy{\UU}{\sigma}            & \reduce \UU \\
  \subTy{(\El\, u)}{\sigma}      & \reduce \El\,(\subTm{u}{\sigma}) \\
  \subTy{(\Pi\,A\,B)}{\sigma}    & \reduce \Pi\,(\subTy{A}{\sigma})\,(\subTy{B}{\sigma\uparrow A}) \\
  \subTy{(\mathsf{Id}\;a\;t\;u)}{\sigma} & \reduce \mathsf{Id}\,(\subTm{a}{\sigma})\,(\subTm{t}{\sigma})\,(\subTm{u}{\sigma})
\end{alignat*}

\begin{alignat*}{5}
  ;\text{-}\mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
  ;\text{-}\mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , \sub{t}{\sigma} \\
  \emptyctx\eta   & : \implicit{\Gamma, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\;\Gamma,\Delta} &&  \sigma
\end{alignat*}

\begin{alignat*}{5}
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\;\Gamma\; A} &&  t \\
  [\idS]         & : \implicit{\Gamma, A, t}\; && {\sub{t}{\idS}}          && =^{\Tm\;\Gamma\;A} && t \\
  [;]            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {\sub{t}{\sigma ; \tau}} && =^{\Tm\;\Gamma\;\subTy{A}{\sigma ; \tau}} && {\sub{(\sub{t}{\tau})}{\sigma}} \\
  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& \sub{(\lambda\,t)}{\sigma} && {} =^{\Tm\,\Gamma\,(\subTy{(\Pi\,A\,B)}{\sigma})} {} && \lambda (\sub{t}{\sigma \uparrow A}) \\
  \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\,(\Gamma, A)\,B} && t \\
  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\,\Gamma\,(\Pi\,A\,B)} && \lambda(\mathsf{app}\,t) \\
  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& \sub{(\mathsf{refl}\;t)}{\sigma} && =^{\Tm\;\Gamma\;(\subTy{(\mathsf{Id}\,a\,t\,t)}{\sigma})} {} && \mathsf{refl}(\sub{t}{\sigma})
\end{alignat*}

\begin{alignat*}{5}
  \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\mathsf{Id}\,a\,t\,u) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
\end{alignat*}
